 Quantile Regression Analysis for Statin Effects on Body Mass Index

### 2021-09-17 manuscript/manuscript2.Rmd
Update

 
---
title: " Quantile Regression Analysis for Statin Effects on Body mass Index"
output:
    bookdown::pdf_document2:  
        keep_tex: true
bibliography: cite.bib
link_citations: true
header-includes:
    - \usepackage{setspace}\doublespacing 
    - \usepackage{float}
    - \floatplacement{figure}{H}
abstract: "Body Mass index plays an important role in predicting cardiovascular  diseases, strokes, diabetes and many others. In this article we use quantile regression to study the effects of age, total cholesterol, cholesterol drug use on the conditional distribution of BMI and it is been compared to the result obtained by ordinary least squares approach. We study age and total cholesterol effect on the BMI in two different ways: quadratic terms and spline basis expansions. The national health and nutrition examination survey (NHANES) data is used in this study. The differences between using quadratic terms and spline basis expansion for the predictors is also presented. "
keywords: "My keywords"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 12pt
fig_caption: yes
indent: true
---
<!-- \setstretch{4} -->
\newpage
\section{Introduction}

  High body mass index (BMI) has a negative impact on public health[@nuttall2015body]. There is a positive relation between the risk of death from all causes, cardiovascular disease, cancer, or other diseases and BMI for both genders[@calle1999body]. Due to high cost of obesity, it is crucial to get better understanding of the underlying risk factors for obesity. This can help decision makers to create  new recommendations that help to prevent and stop the dramatic increase in BMI.

  BMI plays an important role in predicting heart disease risk [@katbody]. Approximately 18 million deaths annually per year world wide are caused by cardiovascular diseases(CVD), and this number is similar to the number of nonfatal cardiovascular events [@hay2017gbd].
In 2011, annual costs for CVD and stroke were estimated at $\$320.1$ billion, which is more than spending on cancer. The total CVD cost includes \$195.6  billion in direct costs (health-care costs) and \$124.5 billion of future productivity loss [@mozaffarian2015executive]. Important risk factors associated with CVD are high BMI and abnormal lipid ratio  [@yusuf2004effect;@anderson1991]. There are some drugs that are used reduce CVD risks by lowering low-density lipoprotein (LDL) like statin.  Statin use  reduces the risk of cardiovascular diseases, even with a population with no CVD risks  [@yusuf2016]. On the other hand, statin uses associate with negative side effects on health conditions, for example, BMI. The effect of BMI on the choice of lipid-lowering treatment has been studied by [@ferrieres2018]. They found that there is a positive  correlation ($\rho=0.13$) between statin does levels and  BMI levels.
Additionally, taking cholesterol drug medication  associates with a high risk of elevating BMI level. It is been shown that statin users consume 192 additional calories per day which causes gaining a 6 lb to 11 lb. in a year. Statin users gain 1.3 units in the BMI measure while non-staining users gain 0.4 $kg/m^2$. Moreover, consumption of fat in statin users raised by 14.4%[@sugiyama2014]. This study gives us insight about the association between cholesterol drug uses and condition means of the BMI.

An important question to ask is, are there  an associations between cholesterol drug uses and different BMI quantiles? This question helps us to understand cholesterol drug uses effect on the  obese, overweight population, so that a decision maker take an optimal action about treatment policy.

  Ordinary least squares (OLS) regression helps us to study the effects of predictors on the mean of the response. For example, estimating the association of different factors like a lipid-lowering drug with the conditional mean of BMI [@ferrieres2018]. OLS approaches does not give us insight into the predictor's effects on different quantiles of the responses.    
  However, quantile regression (QR) is used to investigate the heterogeneity in the association of the $\tau$th quantile $0<\tau<1$ of BMI with a set of independent predictors to investigate the effects of a specific predictors on the various quantiles of the BMI. Estimation of underweight and overweight regressed on age could be hard because of the trend change and nonlinearity behavior of the BMI with respect to age, as it is shown in [@flegal1999].
  <!-- BMI departs normality because of the skewness on the right and left tails. -->
  Therefore, QR is an appropriate method used to estimate the differences in the  association of the BMI with predictors.

@@

 We compare two approaches to study the QR of BMI regressed on age and total cholesterol (TC) levels.  One approach is called quantile polynomial regression. It is defined as a way to model the non-relationship between predictors and distribution of the  response as an $n$th degree polynomial in the predictors. And the other approach is using a spline basis expansion of the predictors.   
  Polynomial regression using the quadratic term of predictors forces the response to take convex or concave shape, see Figure 1.12 [@koenker2005]. This is because the limit of the response variable $y$, when modeled as a quadratic function of the predictor variable $x$,  approach $\pm\infty$ ( i.e., as $x \rightarrow\pm\infty,$ $y\rightarrow\pm\infty).$ However, when we model the response variable using a spline basis expansion of the predictor variable the response is not forced to take a specific shape. In the latter case, different polynomials are constructed for the different regions in the range of predictors that satisfy continuity and smoothness conditions at the knots. 
  Because of the above issue in the polynomial regression, we found that spline regressions produce a different estimate than polynomial regressions.

 \section{Quantile Regression}
QR is a tool used to regress the dependent variable with high variance over the independent variables. QR is developed to study the relationships between predictors and the response when characteristics other than conditional mean of the response are of interest. Also, QR is a valuable tool to study association between predictors and response when we have a week or no-relationships between predictors and the conditional mean of the response. Moreover, one of the advantages of using QR over OLS is QR is robust for outliers.
A brief description of quantile regression is introduced in the following paragraphs.
For a continuous random variable $X$, the cumulative distribution function (CDF) is $$F(X)=P(X\leq x),$$ and the $\tau$th quantile of $X$ is defined by $$ F^{-1}(\tau)=\text{inf}\{x: F(x)\ge \tau\}, $$ where $0<\tau<1$.
Let the loss function of interest be defined by $$\rho_{\tau}(u)= u(\tau-I_{(u<0)}),$$ where $I$ is the indicator function  [@koenker2005]. The quantile estimator is the value that minimizes the expected loss function

@@

OLS method expresses the conditional mean of $y$ given $x$ as $\mu(x)=x^T\beta$ and it solves $$ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n(y_i- x_i^T\beta)^2.$$ 
The above equation represent the loss function for the OLS regression.By solving this equation for the minimum values for $\beta,$ we get regression equation, which is $\hat{y}=x_i^T\beta$.
On the other hand, QR expresses the conditional quantile function $Q_y(\tau|x)=x^T \beta ({\tau)}$ and solves $$ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n\rho_{\tau}(y_i- x_i^T\beta)^2.$$
The above This minimization problem can be reformulated to a linear programming problem and it can be solved using standard statistical software.

@@

\section{Splines Verses Polynomials}
    Most of the relations between the responses and predictors variables are complicated to the point that linear regressions are not suitable to model these relationships[@bruce2020]. For example, the response to different levels of drug doses is not a linear relationship. Linear regressions can be generalized to deal with nonlinear effects. One approach is through including polynomial terms in the regression equation. A continuous predictor can be modeled as a linear, say $x$, or nonlinear term, say $x^2$ depending on the relationship with the response variable.
    The mathematical model for $n$ degree polynomial regression is shown in the Eq\eqref{non}
 \begin{equation}\label{non}
 y=\beta_0+\beta_1 X+\beta_2 X^2+\cdots+\beta_nX^n.
 \end{equation}

 One of the limitations of using polynomial regression is the curvature that can be captured is limited with low order terms. However, including higher-order terms has a negative impact on the model by introducing undesirable "wiggliness" in the regression equation especially near the boundaries of the data.  Another robust approach to model nonlinear relationships is splines. It is similar to a technique used by draftsmen in plotting curves. 
 The spline is a process of constructing a set of piece-wise continuous polynomials that are smoothly connected at a set of points in the range of the predictor variable. The connection points are called knots i.e. splines are used to smoothly interpolate between certain points. Let $a,b$ ,and $c$ are the endpoints of the $x-$axis intervals.
 A transformation for some or all of predictors is needed to capture the nonlinearity in the model. The family of transformation of the predictors that can be fit together to built the model's shape is known as a basis function. 
the basis functions are $b_1(x),b_2(x), \cdots,b_k(x),$ and the estimation of $y_i$ is computed as follows:

@@

\section{Methods}
A multivariate quantile regression model is used to assess the characteristics of the association variability in different quantiles of the conditional distribution of the body mass index.

 The independent variables in our model are gender, race, age, total cholesterol and cholesterol drug use (yes or no). All types of cholesterol drugs are used including statins. The included races are non-Hispanic white, non-Hispanic black, Hispanic, and others. The relation between total cholesterol and fasting glucose has been studied in [@tsaousis2014; @chang2011]. They found that there is a positive correlation between the two groups on average. Therefore, fasting glucose level is not included to avoid collinearity. The following formulas represents the different QR models that fit conditional distribution of BMI on :

 \begin{itemize}
 \item Model I: Age enters the model as spline basis expansion, Race, Gender, and cholesterol drug use (Yes or No).
 \item Model II: TC enters the model as spline basis expansion, Race, Gender, and cholesterol drug use.
 \item Model III: Age enters the model as a quadratic term, Race, Gender, and cholesterol drug use.
 \item Model IV: TC enters the model as a quadratic term, Race, Gender, and cholesterol drug use. 
 \end{itemize}

@@

\section{Numerical Example}
\subsection{Data}
 The data used in this study is the National Health and Nutrition Examination Survey data (NHANES)[@NHANES]. The survey examines a nationally representative sample of the U.S. population focusing on a variety of health and nutrition measurements released in two year cycle.  In this study, we accumulated 6 cycles of NHANES data (2007–2018). We used two data files: One contains demographic variables, such as age, sex, race, income, etc. and the other contains data that are related to body measurements, such as BMI, head circumference, etc. These files are merged by using the respondent sequence number (SEQN) resulting in a around 12,000 records. The missing values are deleted. We selected a population age between 20 and 80.
 BMI are classified into different categories according to underweight, 18.5 $kg/m^2$; normal weight, 18.5 to 25 $kg/m^2$; overweight, 25 to 30 $kg/m^2$; obese, 30 to 35 $kg/m^2$; and very obese more than 35 $kg/m^2$.

| Syntax      | Male  | Female     |
| :---        |    :----:   |          ---: |
| count |  5990      | 6416   |
| Mean of Age | 49.9       | 49.73   |
| BMI        |  28.549    |  29.379  |
| TC         |        |    |
| Statin use (ratio) | 0.198       | 0.181   |

```{r box, out.width = "100%", fig.align = "center", fig.cap = "Illustration of data distribution using a box plot. The plot compares cholesterol drug users vs non-cholesterol drug users", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "base_box_plot.png"))) {
    source(here::here("exploratory-data-analysis.R"))
}
knitr::include_graphics(here::here("images", "base_box_plot.png"))
```

Figure \@ref(fig:box) shows the association between age and BMI with respect to cholesterol drug changes based on age. At the younger ages, the correlation is positive while at the middle age the correlation is almost flat, and at the old age the correlation is negative. Moreover, at low age cholesterol medications are  described only for a population with high BMI, but at a higher age, the difference in term of the BMI for a population that takes cholesterol medication and without cholesterol medication decreases until it diminishes as in the white population. The  heterogeneity can also be seen in gender ?, for the Hispanic female population, the cholesterol medication is taken by the population with lower BMI if compared to the population that does not take cholesterol medication.

@@

From the OLS it is obvious cholesterol drug users have on average higher BMI levels if compared to non-cholesterol drug users which are around 1.75. The disparity is almost regular through different BMI quantiles. For example, in the lower quantiles of the distribution the cholesterol drug users have higher BMI value by about 1.5 $kg/m^2$, but around at 50 th percentile of the conditional distribution the difference is 1.8 $kg/m^2$. Overall,  cholesterol drug use seems to be associated with rather large effects on BMI levels somewhat between  1 to 2 $kg/m^2$.

@@

Figure \@ref(fig:resu1000) illustrate the differences in modeling Age in two different ways (Model I; spline basis expansion) and (Model III; quadratic term ).  
 Model I shows, at the lower quantile, the marginal age effects on BMI increases from age 20 to around age 42, and then it stays flat up to age 70 but it starts to decrease after that. While model III shows that there is positive relationship between Age and BMI in the age range between 20 and 50. Then age effects stay flat in the age range between 50 and 60 years old and decreases after that.

 At the 50th quantile, model I shows the highest age effect on BMI occurs earlier by 10 years when compares to the lower quantile, but model III shows maximum effect occurs at around age 50, which is similar to the age effect at lower quantiles. The age effects start to decrease using Model I at around 70 years old while model III shows the decay starts at  around age 60.   
 At higher quantiles $\tau=75$, using Model I, the maximum age effects occurs earlier by three years, but Model III shows the highest age effects occurs at around age 50. The decay, according to Model I, starts at around age 70 years old, but Model III shows the decay starts earlier at around 70 years old. Convex behavior of the quadratic term has a clear pattern in formulating age effects on the BMI.  The box plot (Fig. \@ref(fig:box))   supports Model I results because the highest mean and inter quantile range occurs at age 30.

Using Model I, at the 90th quantile of the conditional BMI distribution, we have a similar behavior as in 0.75th quantile except the relationship is stronger, i.e, for example, the age effects at age 20 is 37 $kg/m^2$ while at the later the effect is 30 $kg/m^2$. The BMI trend in this modeling is close to the trend shown in [@chen2005], which computed using a complicated polynomial and log transformation for the response. Our conclusion from Figure \@ref(fig:resu1000) is that the marginal age effects on BMI is different on different quantiles. Moreover, modeling age as quadratic term or as spline basis expansion produce different effect with respect to different  quantiles of the conditional BMI distribution.

```{r resu1000, out.width = "100%", fig.align = "center", fig.cap = "  An illustration for the differences in modeling the age effects on different BMI quantiles. The predictors (Age) modeled using quadratic terms and basis splines expension.", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "comp_age_mixed")))  
knitr::include_graphics(here::here("images", "comp_age_mixed.png"))
```
\subsection{TC effects Analysis}
Next, we investigate marginal effect of TC on four different quantiles of the conditional distribution of the BMI. 
The quadratic effect of total cholesterol on the conditional distribution of BMI is convex in general when age is modeled using quadratic term. However, age effect on the BMI is different when age is modeled using spline basis expansion. The disparity is due to the same reason mentioned earlier, which is as
TC$\rightarrow\pm\infty,$ BMI$\rightarrow 0).$ At the lower tail ($\tau=0.1$), for TC less than 300 mg$\setminus$dl, we have positive association, that is, when TC increases, the BMI increases also. Moreover, when TC passes this level, spline basis expansion modeling of TC presents negative correlation between TC and BMI while quadratic modeling of TC shows positive correlation.

At the 50th quantile of BMI, the two methods of modeling TC shows almost the same trend through whole TC range.   
 At higher quantiles $(\tau=75)$, for TC values less than 350, the association is negative that is as TC cholesterol level increases the BMI decreases, for both ways of modeling TC. However, the two models differ after TC passes this value. The quadratic term force the TC effect to take convex shape.
 At a higher quantile $(\tau=90),$  both ways of modeling TC shows BMI take a convex shape, but quadratic way of modeling BMI is more convex than the other. When TC passes 380 mg$\setminus$dL, where TC modeled using spline basis expansion, the BMI starts to decrease slow if compared to the quadratic way for modeling BMI, see  Figure \@ref(fig:resu1010).

 <!-- On the other hand, when we use splines to modeled the total cholesterol, there is a huge difference in the shape of the association expect at the higher quantile there is some similarity. At the lower quantile, there is negative correlation between total cholesterol and BMI for total cholesterol in the range less than 160. Then the correlation becomes positive for total cholesterol in the range of (170, 250). At the 55th quantile, the correlation is positive for low cholesterol level and then becomes almost flat for the rest cholesterol values. -->

 The population with low BMI, it is expected when TC increases (more than 200 mg$\setminus$dL), people start to reduce their BMI. However, this trend is captured by using spline basis expansion for the age but it was missed by using quadratic term for the age, see the upper left corner plot in Figure \@ref(fig:resu1010). In addition, he$/$she with high BMI reduces their weight to reduce him$/$her TC. The quadratic modeling of the TC shows a sharp decrease in the BMI if compared to spline basis expansion. This can be seen in the lower right corner of Figure \@ref(fig:resu1010). As it was shown in [@mc2020effects] the obese population has higher TC, this trend is not captured by quadratic term modeling for TC.

```{r resu1010, out.width = "100%", fig.align = "center", fig.cap = " An illustration of  marginal effects of total cholesterol  on different BMI quantiles. The predictors modeled using quadratic terms and splines. The BMI quantiles $\tau= 0.1, 0.5, 0.75, 0.90$ coressponds to  21.5,  27.9,  32.4, 37.7 $kg/m^2$, respectively. ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "comp_tot_mixed")))  
knitr::include_graphics(here::here("images", "comp_tot_mixed.png"))
```

\subsection{Analysing Race$\setminus$ Ethinicity effect on BMI}

>>>>

 
---
title: " Quantile Regression Analysis for Statin Effects on Body Mass Index"
output:
    bookdown::pdf_document2:  
        keep_tex: true
bibliography: cite.bib
link_citations: true
header-includes:
    - \usepackage{setspace}\doublespacing 
    - \usepackage{float}
    - \floatplacement{figure}{H}
abstract: " An individuals body mass index (BMI) is an important predictor of a  cardiovascular  diseases, stroke, diabetes and many other medical conditions. In this article, we use quantile regression to study the effects of age, total cholesterol, and use of medication on the conditional distribution of BMI and compare these results to those obtained by ordinary least squares. We study age and total cholesterol effect on the BMI in two different ways: quadratic terms and spline basis expansions. The national health and nutrition examination survey (NHANES) data is used in this study. The differences between using quadratic terms and spline basis expansion for the predictors is also presented. "
keywords: "My keywords"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 12pt
fig_caption: yes
indent: true
---
<!-- \setstretch{4} -->
\newpage
\section{Introduction}

  High body mass index (BMI) has a negative impact on an individuals health [@nuttall2015body] because there is a positive relationship between the risk of death from all causes including cardiovascular disease, cancer, or other diseases and BMI for both genders [@calle1999body]. Due to high cost of obesity on an individuals health, it is crucial to get better understanding of the underlying risk factors for obesity. Understanding the causes of obesity can help decision makers  create  new recommendations that help to prevent and stop the dramatic increase in BMI.

  BMI plays an important role in predicting heart disease risk [@katbody]. Approximately 18 million deaths annually per year world wide are caused by cardiovascular diseases(CVD), and this number is similar to the number of nonfatal cardiovascular events [@hay2017gbd].
In 2011, annual costs for CVD and stroke were estimated at $\$320.1$ billion, which is more than what is spent on cancer. The total CVD cost includes \$195.6  billion in direct costs (health-care costs) and \$124.5 billion of future productivity loss [@mozaffarian2015executive]. Important risk factors associated with CVD are high BMI and abnormal lipid ratio  [@yusuf2004effect;@anderson1991]. There are some medications that are used reduce CVD risks by lowering low-density lipoprotein (LDL) including statins.  Statin use  reduces the risk of cardiovascular diseases, even with a population with no CVD risk factors [@yusuf2016]. On the other hand, statin uses is associate with negative side effects on health conditions including increased BMI. The effect of BMI on the choice of lipid-lowering treatment has been studied by [@ferrieres2018]. Where they found that there is a positive  correlation ($\rho=0.13$) between statin does levels and  BMI levels.
Additionally, it is been shown that statin users consume on average 192 additional calories per day which causes gaining on average  a 6 to 11 lbs in a year. Statin users gain on average 1.3 units in the BMI measure while non-staining users gain on average 0.4 $kg/m^2$. Moreover, consumption of fat in statin users raised on average by 14.4%[@sugiyama2014]. This study gives us insight about the association between cholesterol medication uses and conditional means of the BMI.

An important question to ask is, are there  an associations between cholesterol medication uses and different BMI quantiles? This question helps us to understand cholesterol medication uses effect on the  obese, overweight population, so that a decision maker take an optimal action about treatment policy.

  Ordinary least squares (OLS) regression helps us to study the effects of predictors on the mean of the response. For example, OLS is used to estimate the conditional mean of BMI given different factors like a lipid-lowering medications [@ferrieres2018]. However, OLS approaches does not give us insight into the predictor's effects on different quantiles of the responses.    
  Instead, quantile regression (QR) is used to investigate the conditional relationship heterogeneity of the $\tau$th quantile $0<\tau<1$ of BMI given with a set of independent predictors.
  <!-- to investigate the effects of a specific predictors on the various quantiles of the BMI. -->

  <!-- BMI departs normality because of the skewness on the right and left tails. -->

@@

Estimating the average age effects on BMI could be hard because the age effects is varying with respect to BMI levels. In addition, it is been found that the  BMI behaves nonlinear with respect to age [@flegal1999].
 <!-- Therefore, QR is an appropriate method used to estimate the differences in the  association of the BMI with predictors.  -->
   We compare two approaches to study the QR of BMI regressed on age and total cholesterol (TC) levels.  One approach is to model the conditional distribution of the BMI as a polynomial function of the age, which is called quantile polynomial regression.A quantile polynomial regression is defined as a way to model the non-linear relationship between predictors and distribution of the  response as an $n$th degree polynomial in the predictors. And the other approach is using a spline basis expansion of the predictors.   
  Polynomial regression using the quadratic term of predictors forces the response to take convex or concave shape, see Figure 1.12 [@koenker2005]. This is because the limit of the response variable $y$, when modeled as a quadratic function of the predictor variable $x$,  approach $\pm\infty$ ( i.e., as $x \rightarrow\pm\infty,$ $y\rightarrow\pm\infty).$ However, when we model the response variable using a spline basis expansion of the predictor variable the response is not forced to take a specific shape. In the latter case, different polynomials are constructed for the different regions in the range of predictors that satisfy continuity and smoothness conditions at the knots. 
  Because of the above issue in the polynomial regression, we found that spline regressions produce a different estimate than polynomial regressions.

 \section{Quantile Regression}
QR is a tool used to regress the dependent variable with high variance over the independent variables. QR is developed to study the relationships between predictors and the response when characteristics other than conditional mean of the response are of interest. Also, QR is a valuable tool to study association between predictors and response when we have a week or no-relationships between predictors and the conditional mean of the response. Moreover, one of the advantages of using QR over OLS is QR is robust to outliers.
A brief description of quantile regression is introduced in the following paragraphs.
For a continuous random variable $X$, we denote by $x$ for an instances of $X$, the cumulative distribution function (CDF) is $$F(X)=P(X\leq x),$$ and the $\tau$th quantile of $X$ is defined by $$ F^{-1}(\tau)=\text{inf}\{x: F(x)\ge \tau\}, $$ where $0<\tau<1$.
Let the loss function of interest be defined by $$\rho_{\tau}(u)= u(\tau-I_{(u<0)}),$$ where $I_{\cdots}$ is the indicator function  [@koenker2005]. The quantile estimator is the value that minimizes the expected loss function

@@

OLS method expresses the conditional mean of $y$ given $x$ as $\mu(x)=x^T\beta$ and it solves $$ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n(y_i- x_i^T\beta)^2.$$ 
The above equation represent a function of the loss for the OLS regression. By solving this equation for $\beta,$ we get the estimated  regression equation  $\hat{y}=x_i^T\beta$.
On the other hand, QR expresses the conditional quantile function $Q_y(\tau|x)=x^T \beta ({\tau)}$ and solves $$ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n\rho_{\tau}(y_i- x_i^T\beta)^2.$$
The above minimization problem can be reformulated to a linear programming problem and it can be solved using standard statistical software.

@@

\section{Accounting for a Nonlinear Functional Response}
    Most of the relationships between the responses and predictors variables are nonlinear so that linear regressions are not suitable to model these relationships [@bruce2020]. For example, the response to different levels of medication doses is not a linear relationship. Linear regressions can be generalized to deal with nonlinear responses in a variety of ways. One approach is through including polynomial terms in the regression equation. A continuous predictor can be modeled as a linear term (i.e.,$x$) or nonlinear term (i.e., $x^2$) depending on the relationship with the response variable.
    The mathematical model for $n$ degree polynomial regression is shown in the Eq\eqref{non}
 \begin{equation}\label{non}
 y=\beta_0+\beta_1 x+\beta_2 x^2+\cdots+\beta_n x^n+\epsilon.
 \end{equation}

 One of the limitations of using polynomial regression is the curvature that can be captured is limited with low order terms. However, including higher-order terms has a negative impact on the model by introducing undesirable "wiggliness" in the regression equation, especially near the boundaries of the data.  Another robust approach to model nonlinear relationships is splines, which is similar to a technique used by draftsmen in plotting curves. 
 The spline is a process of constructing a set of piece-wise continuous polynomials that are smoothly connected at a set of points called knots in the range of the predictor variable and are used to smoothly interpolate between points. The family of transformation of the predictors that can be fit together to built the model's shape is known as a basis function. 
the basis functions are $b_1(x),b_2(x), \cdots,b_k(x),$ and the estimation of $y_i$ is computed as follows:

@@

\section{Methods}
A multivariate quantile regression model is used to assess the characteristics of the association variability in different quantiles of the conditional distribution of the BMI.

 The independent variables in our model are gender, race, age, total cholesterol and cholesterol medication(yes or no). The "yes" response for the variable "cholesterol medication" refers to using any types of cholesterol medication including statin. The included races/ethnicity are non-Hispanic white, non-Hispanic black, Hispanic, and others. The relation between total cholesterol and fasting glucose was been studied in [@tsaousis2014; @chang2011]. They found that there is a positive correlation between total cholesterol and fasting glucose on average. Therefore, fasting glucose level is not included in our analysis to avoid collinearity. The following formulas represents the different QR models that fit conditional distribution of BMI on :

 \begin{itemize}
 \item Model I: \textit{Age enters the model as a quadratic term, Race, Gender, and cholesterol medication use.}
 \item Model II:\textit{TC enters the model as a quadratic term, Race, Gender, and cholesterol medication use.}
 \item Model III:\textit{Age enters the model as spline basis expansion, Race, Gender, and cholesterol medication use (Yes or No).}
 \item Model IV: \textit{TC enters the model as spline basis expansion, Race, Gender, and cholesterol medication use.}
 \end{itemize}

@@

\section{Numerical Example}
\subsection{Data}
 The data used in this study is from the National Health and Nutrition Examination Survey (NHANES, @NHANES). The survey examines a nationally representative sample of the U.S. population focusing on a variety of health and nutrition measurements released in two year cycle.  In this study, we accumulated 6 cycles of NHANES data (2007–2018). We used two data files: One contains demographic variables, such as age, sex, race, income, etc. and the other contains data that are related to body measurements, such as BMI, head circumference, etc. These files are merged by using the respondent sequence number (SEQN), resulting in  around 12,000 records after missing values are deleted. We selected an adult population aged between 20 and 80.  On average, adults female population  have higher  BMI. The race/ethnicity variables denotes for Mexican American, Other Hispanic, Non-Hispanic White, Non-Hispanic Black, and Other Race - Including Multi-Racial.
 BMI are classified into different categories according to underweight, 18.5 $kg/m^2$; normal weight, 18.5 to 25 $kg/m^2$; overweight, 25 to 30 $kg/m^2$; obese, 30 to 35 $kg/m^2$; and very obese more than 35 $kg/m^2$.

|       | Male  | Female     |
| :---        |    :----:   |          ---: |
| count |  5990      | 6416   |
| Mean of Age | 49.9       | 49.73   |
|  Mean of BMI        |  28.549    |  29.379  |
| Mean of TC         |  189.641      |  195.977  |
| Statin use (ratio) | 0.198       | 0.181   |

```{r box, out.width = "100%", fig.align = "center", fig.cap = "Illustration of data distribution using a box plot. The plot compares cholesterol medicated users vs non-cholesterol medicated users", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "base_box_plot.png"))) {
    source(here::here("exploratory-data-analysis.R"))
}
knitr::include_graphics(here::here("images", "base_box_plot.png"))
```

Figure \@ref(fig:box) shows how the association between age and BMI given cholesterol medication use changes based on age. At younger ages, the average BMI for black population increases as a function of age up to around age 30 years. For ages more than 40 years, the association becomes negative. This observation suggests that at around 30 year old people start to take care of their weights. Mexican Females  do not show significant reduction in the BMI as a function of age. For White Males, the association is flat at the middle ages and negative at older ages. Moreover, at lower ages, cholesterol medications are  prescribed only for a population with high varying BMI, but at  older ages, the difference in BMI given cholesterol medication status decreases until it diminishes as in the white population. The  heterogeneity can also be seen in gender, for the Hispanic female population, the cholesterol medication is taken by the population with lower BMI if compared to the population that does not take cholesterol medication.

@@

From the OLS it is obvious cholesterol medication users have on average higher BMI levels if compared to non-cholesterol medicated users which are around 1.75. The disparity is almost consistent throughout different BMI quantiles. For example, in the lower quantiles of the distribution the cholesterol medicated users have higher BMI value by about 1.5 $kg/m^2$ on average, but around at 50 th percentile of the conditional distribution the difference is on average 1.8 $kg/m^2$. Overall,  cholesterol medication use seems to be associated with rather large effects on BMI levels somewhat between  1 to 2 $kg/m^2$ on average.

@@

Figure \@ref(fig:resu1000) illustrate the differences in modeling age in two different ways (Model III; spline basis expansion) and (Model I; quadratic term ).  
 Model III shows, at the lower quantile, the marginal age effects on BMI increases from age 20 to around age 42, and then it stays flat up to age 70 but it starts to decrease after that. While model I shows that there is positive relationship between Age and BMI in the age range between 20 and 50. Then age effects stay flat in the age range between 50 and 60 years old and decreases after that.

 At the 50th quantile, model III shows the highest age effect on BMI occurs earlier by 10 years when compares to the lower quantile, but model I shows maximum effect occurs at around age 50, which is similar to the age effect at lower quantiles. The age effects start to decrease using Model III at around 70 years old while model I shows the decay starts at  around age 60.   
 At higher quantiles ($\tau=75$) Model III shows the maximum age effects occurs at age around 32 years old, which is  earlier than the age at which the peak coccurs in  by three years than , but Model I shows the highest age effects occurs at around age 50. The decay, according to Model III, starts at around age 70 years old, but Model I shows the decay starts earlier at around 70 years old. Convex behavior of the quadratic term has a clear pattern in formulating age effects on the BMI.  The box plot (Fig. \@ref(fig:box))   supports Model III results because the highest mean and inter quantile range occurs at age 30.

Using Model III, at the 90th quantile of the conditional BMI distribution, we have a similar behavior as in 0.75th quantile except the relationship is stronger, i.e, for example, the age effects at age 20 is 37 $kg/m^2$ while at the later the effect is 30 $kg/m^2$. The BMI trend in this modeling is close to the trend shown in [@chen2005], which computed using a complicated polynomial and log transformation for the response. Our conclusion from Figure \@ref(fig:resu1000) is that the marginal age effects on BMI is different on different quantiles. Moreover, modeling age as quadratic term or as spline basis expansion produce different effect with respect to different  quantiles of the conditional BMI distribution.

```{r resu1000, out.width = "100%", fig.align = "center", fig.cap = "  An illustration for the differences in modeling the age effects on different BMI quantiles. The predictors (Age) modeled using quadratic terms and basis splines expension.", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "comp_age_mixed")))  
knitr::include_graphics(here::here("images", "comp_age_mixed.png"))
```
\subsection{TC effects Analysis}
Figure \@ref(fig:resu1010) presents the marginal effect of TC for four different quantiles of the conditional distribution of the BMI. 
The quadratic effect of total cholesterol on the conditional distribution of BMI is required to be convex when age is modeled using quadratic term. However, TC effect on the BMI is different when TC is modeled using spline basis expansion. The disparity is due to the same reason mentioned earlier, which is as
TC$\rightarrow\pm\infty,$ BMI$\rightarrow \pm\infty).$ At the lower tail ($\tau=0.1$), for TC less than 300 mg$\setminus$dl, we have positive association, that is, when TC increases, the BMI increases also, on average. Moreover, when TC passes 300 mg$\setminus$dl level, spline basis expansion modeling of TC presents negative correlation between TC and BMI while quadratic modeling of TC shows positive correlation. In addition, the confidence interval (CI) for quadratic modeling of age is larger than the CI for spline basis expansion modeling for the age. The CI for quadratic modeling of the age indicate poor fitting while spline basis expansion perform better.

At the 50th quantile of BMI, the two methods of modeling TC shows almost the same trend through whole TC range.   
 At higher quantiles $(\tau=75)$, for TC values less than 350 mg/dL, the association is negative that is as TC cholesterol level increases the BMI decreases, for both functional form for modeling TC. However, the two models differ after TC values passes 350 mg/dL because quadratic functional form forces the TC effect to take a convex shape.
 At a higher quantile $(\tau=90),$  both the models for TC show BMI taking a convex shape, but quadratic way of modeling BMI is more parabolic than the spline model. When TC passes 380 mg$\setminus$dL, the TC model using the spline basis expansion shows the BMI starts to decrease more slowly  compared to the quadratic.

 ```{r resu1010, out.width = "100%", fig.align = "center", fig.cap = " An illustration of  marginal effects of total cholesterol  on different BMI quantiles. The predictors modeled using quadratic terms and splines. The BMI quantiles $\tau=$ 0.1, 0.5, 0.75 and 0.90 coressponds to  21.5,  27.9,  32.4 and 37.7 $kg/m^2$, respectively. ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "comp_tot_mixed")))  
knitr::include_graphics(here::here("images", "comp_tot_mixed.png"))
```
 <!-- On the other hand, when we use splines to modeled the total cholesterol, there is a huge difference in the shape of the association expect at the higher quantile there is some similarity. At the lower quantile, there is negative correlation between total cholesterol and BMI for total cholesterol in the range less than 160. Then the correlation becomes positive for total cholesterol in the range of (170, 250). At the 55th quantile, the correlation is positive for low cholesterol level and then becomes almost flat for the rest cholesterol values. -->

 For a population with low BMI, it is expected when TC increases (more than 200 mg$\setminus$dL), individuals start to reduce their BMI, on average. However, this trend is captured by using spline basis expansion for the TC but it was missed by using quadratic model for TC ( upper left corner plot in Figure \@ref(fig:resu1010).) In addition, an individual with high BMI reduces their weight to reduce their TC. The quadratic modeling of the TC shows a sharp decrease in the BMI when compared to spline basis expansion at $\tau=0.9$ ( lower right corner of Figure \@ref(fig:resu1010).) As it was shown in [@mc2020effects] the obese population has a higher TC and this trend is not captured by quadratic term modeling for TC.

\subsection{Analysing Race/Ethinicity effect on BMI}

Figure \@ref(fig:resu3) illustrates age effects on 90th BMI quantile, where age modeled as a quadratic term. There is a clear association between age and BMI. In general,  age effects on BMI, considering race ethnicity, behaves at a similar way as the whole population, which is a parabolic shape. For example, a young Mexican female not using cholesterol medication has an average BMI around 40 $kg/m^2$, but the BMI values decreases later up to 35 $kg/m^2$ at age 80 years. The effects of age is higher for cholesterol medicated users by an average of about about 2 $kg/m^2$. Moreover,  Mexican males have a higher average BMI  than females by about 2 $kg/m^2$ at $\tau =0.9$. The lowest age effects on 90th BMI quantile is found in the white race.
In general age effect on BMI for males are higher than for females.  Figure \@ref(fig:resu33) presents age effect on BMI through using spline basis expansion.
